---
title: "01 data cohort cleaning"
author: "Ruiyi Tian"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
source("R/functions.R")
load_core_packages()
```

# Purpose

This file does all **upstream** work:

1. import raw data
2. harmonize variables
3. apply manuscript-aligned inclusion and exclusion criteria
4. derive biological aging metrics
5. save frozen analysis-ready datasets for downstream analyses

The goal is that **all later Rmd files only read derived datasets created here**.

# 1. Configuration

```{r config}
# Update these to your machine / repository structure
path_raw_ukbb   <- "data_raw/ukbb/"
path_derived    <- "data_derived/"
path_results    <- "results/"

dir.create(path_derived, recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(path_results, "tables"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(path_results, "figures"), recursive = TRUE, showWarnings = FALSE)
```

# 2. Import raw source data

```{r import-data}
# This is a master RData including all analysis file
load("/path/to/Project/Data/Preprocessed_Project_Data.RData") 
```

# 3. Harmonize raw variables

## 3.1 UK Biobank: recode biomarkers needed for clocks

```{r ukbb-recode-biomarkers, eval=FALSE}
# ukbb_raw <- ukbb_raw %>%
#   mutate(
#   # Convert units for each aging biomarkers
#     )), na.rm = TRUE)
#   )
```

## 3.2 Common covariate recode

```{r common-recoding, eval=FALSE}
# Keep this section focused on manuscript variables only.
# Avoid analysis-specific recodes later in downstream files.

# ukbb_raw <- ukbb_raw %>%
#   mutate(
#     # Recategorizing covariates
#     )
#   )

# Female-specific covariates used in female cancer models
ukbb_raw <- ukbb_raw %>%
  mutate(
    ageAtMenarcheCat2 = if_else(ageAtMenarcheCat == "Missing", 13.01432, ageAtMenarche),
    menopauseStatus2 = case_when(
      menopauseStatus == "Missing" & ageBaseline > 51 ~ "Yes",
      menopauseStatus == "Missing" & ageBaseline <= 51 ~ "No",
      menopauseStatus == "Male" ~ "NA",
      TRUE ~ menopauseStatus
    ),
    oralContraceptiveUse2 = case_when(
      oralContraceptiveUse == "Missing" ~ "Past",
      oralContraceptiveUse == "Male" ~ "NA",
      TRUE ~ oralContraceptiveUse
    ),
    parity2 = if_else(parityCat == "Missing", 1.620235, parity)
  )
```

# 4. Manuscript-aligned cohort definitions

The current manuscript uses the following core design choices:

- UK Biobank participants under age 55 at baseline for the main systemic aging analyses
- exclusion of prior cancer except non-melanoma skin cancer
- exclusion of cancer within 6 months of baseline / measurement anchor
- exclusion of underweight participants (`BMI < 18.5`)
- exclusion of missing genetic ancestry for the UK Biobank main models
- clock-specific complete-case biomarker requirements and removal of extreme biomarker values `>5 SD`
- separate analysis-ready datasets for PhenoAge, KDM, metabolomic age, organ aging, and merged downstream files

## 4.1 UK Biobank main analysis cohort

```{r ukbb-main-cohort, eval=FALSE}
ukbb_main_base <- ukbb_raw %>%
  filter(ageBaseline < 55) %>%
  filter(bmi >= 18.5 | is.na(bmi) == FALSE) %>%
  filter(is.na(genetic_ancestry_missing_flag) | genetic_ancestry_missing_flag == 0) %>%
  filter(prior_cancer_excluding_nmsc == 0) %>%
  filter(cancer_within_6m_of_baseline == 0)

```

## 4.2 UK Biobank PhenoAge-ready cohort

```{r ukbb-phenoage-ready, eval=FALSE}
pheno_biomarkers <- c(
  "albumin_gL", "lymph", "mcv", "glucose_mmol", "rdw",
  "creat_umol", "lncrp", "alp", "wbc"
)

ukbb_pheno_ready <- ukbb_main_base %>%
  prepare_phenoage_data(biomarker_cols = pheno_biomarkers, age_col = "age") %>%
  compute_phenoage_from_biomarkers(age_col = "age", biomarker_cols = pheno_biomarkers) %>%
  compute_age_acceleration(age_col = age, ba_col = phenoage, prefix = "phenoage")

```

## 4.3 UK Biobank KDM-ready cohort

```{r ukbb-kdm-ready, eval=FALSE}
kdm_biomarkers <- c(
  "albumin_gL", "alp", "bun", "creat_umol", "lncrp",
  "hba1c", "totchol", "glucose_mmol", "fev", "sbp"
)

ukbb_kdm_ready <- ukbb_main_base %>%
  filter(if_all(all_of(c("age", "sex", kdm_biomarkers)), ~ !is.na(.x))) %>%
  remove_sd_outliers(cols = kdm_biomarkers, threshold = 5) %>%
  compute_kdm_from_biomarkers(sex_col = "sex", biomarker_cols = kdm_biomarkers) %>%
  compute_age_acceleration(age_col = age, ba_col = kdm, prefix = "kdm")

```

## 4.4 Merge main UK Biobank systemic-aging dataset

```{r ukbb-main-merged, eval=FALSE}
ukbb_systemic_main <- ukbb_main_base %>%
  left_join(
    ukbb_pheno_ready %>%
      dplyr::select(eid, phenoage, AA_phenoage, Standardized_AA_phenoage,
                    Tertiles_Standardized_AA_phenoage, Quartiles_Standardized_AA_phenoage),
    by = "eid"
  ) %>%
  left_join(
    ukbb_kdm_ready %>%
      dplyr::select(eid, kdm, AA_kdm, Standardized_AA_kdm,
                    Tertiles_Standardized_AA_kdm, Quartiles_Standardized_AA_kdm),
    by = "eid"
  )

```

## 4.5 Metabolomic and organ-aging analysis datasets

```{r ukbb-met-organ, eval=FALSE}
# Keep these steps here because they belong to upstream dataset creation.
# The exact formulas / imports can remain in source-specific prep scripts if needed,
# but this file should be the single place where the final derived datasets are frozen.

# Example
# ukbb_met_ready   <- ukbb_systemic_main %>% left_join(met_raw, by = "eid")
# ukbb_organ_ready <- ukbb_systemic_main %>% left_join(organ_raw, by = "eid")
```

## 4.6 All of Us validation cohort

```{r aou-ready, eval=FALSE}
### Not publishable due to All of Us policy
```

# 5. Save cohort flow counts used in the manuscript

```{r flow-counts, eval=FALSE}
cohort_flow <- list(
  ukbb_main_base_n = nrow(ukbb_main_base),
  ukbb_pheno_ready_n = nrow(ukbb_pheno_ready),
  ukbb_kdm_ready_n = nrow(ukbb_kdm_ready),
  ukbb_systemic_main_n = nrow(ukbb_systemic_main)
)
```

# 6. Final check

Before moving to `02_main_analysis.Rmd`, confirm that:

- cohort counts match the manuscript
- variable names are harmonized once and only once here
- all downstream scripts can start from `readRDS()` of derived datasets
- no analysis-specific ad hoc cleaning remains outside this file
