---
title: "02 main analysis"
author: "Ruiyi Tian"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
source("R/functions.R")
load_core_packages()
```

# Purpose

This file runs the manuscript's **main analyses only**:

1. birth cohort trend analyses
2. PhenoAge and early-onset cancer associations
3. KDM and early-onset cancer associations
4. metabolomic aging and early-onset cancer associations
5. organ-specific aging and early-onset cancer associations

Everything here should read **frozen** analysis-ready data from `01_data_cohort_cleaning.Rmd`.

# 1. Load derived datasets

```{r load-data}
path_derived <- "data_derived/"
path_results <- "results/"

dir.create(file.path(path_results, "tables"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(path_results, "figures"), recursive = TRUE, showWarnings = FALSE)
dir.create(file.path(path_results, "model_objects"), recursive = TRUE, showWarnings = FALSE)

# Replace with actual file names created in 01.
# ukbb_systemic_main <- readRDS(file.path(path_derived, "ukbb_systemic_main.rds"))
# ukbb_met_ready     <- readRDS(file.path(path_derived, "ukbb_met_ready.rds"))
# ukbb_organ_ready   <- readRDS(file.path(path_derived, "ukbb_organ_ready.rds"))
# aou_systemic_main  <- readRDS(file.path(path_derived, "aou_systemic_main.rds"))

outcomes <- get_manuscript_outcomes()
ukbb_covars <- build_base_covariates("ukbb")
aou_covars  <- build_base_covariates("aou")
```

# 2. Birth cohort trends

## 2.1 Linear birth cohort models

```{r birth-cohort-linear, eval=FALSE}
# UKBB PhenoAge AA
fit_bc_pheno <- fit_birth_cohort_linear(
  data = ukbb_systemic_main,
  outcome = "Standardized_AA_phenoage",
  birth_var = "birthCohort",
  adjust = c("ageBaseline", "ageSquare")
)

# UKBB KDM AA
fit_bc_kdm <- fit_birth_cohort_linear(
  data = ukbb_systemic_main,
  outcome = "Standardized_AA_kdm",
  birth_var = "birthCohort",
  adjust = c("ageBaseline", "ageSquare")
)

# Add metabolomic and AoU analogs here once derived datasets are loaded.
```

## 2.2 GAM birth-year plots

```{r birth-cohort-gam, eval=FALSE}
fit_gam_pheno <- fit_birth_year_gam(
  data = ukbb_systemic_main,
  outcome = "Standardized_AA_phenoage",
  birth_year = "birthYear",
  adjust = c("ageBaseline", "ageSquare")
)

fit_gam_pheno_male <- fit_birth_year_gam(
  data = dplyr::filter(ukbb_systemic_main, sex == "Male"),
  outcome = "Standardized_AA_phenoage",
  birth_year = "birthYear",
  adjust = c("ageBaseline", "ageSquare")
)

fit_gam_pheno_female <- fit_birth_year_gam(
  data = dplyr::filter(ukbb_systemic_main, sex == "Female"),
  outcome = "Standardized_AA_phenoage",
  birth_year = "birthYear",
  adjust = c("ageBaseline", "ageSquare")
)

saveRDS(
  list(overall = fit_gam_pheno, male = fit_gam_pheno_male, female = fit_gam_pheno_female),
  file.path(path_results, "model_objects", "birth_year_gam_pheno.rds")
)
```

# 3. Main Cox models: UK Biobank systemic aging

## 3.1 PhenoAge and early-onset cancer risk

```{r phenoage-main-cox, eval=FALSE}
pheno_cont <- run_cox_series(
  data = dplyr::filter(ukbb_systemic_main, !is.na(Standardized_AA_phenoage)),
  event_vars = outcomes$eo_main_figure,
  time_stop_var = "t_AllCancerEO55_age",
  exposure = "Standardized_AA_phenoage",
  covariates = ukbb_covars,
  time_start_var = "ageBaseline"
)

pheno_cont <- add_bh_fdr(pheno_cont, p_col = "p.value")

pheno_tertiles <- run_cox_series(
  data = dplyr::filter(ukbb_systemic_main, !is.na(Tertiles_Standardized_AA_phenoage)),
  event_vars = outcomes$eo_main_figure,
  time_stop_var = "t_AllCancerEO55_age",
  exposure = "Tertiles_Standardized_AA_phenoage",
  covariates = ukbb_covars,
  time_start_var = "ageBaseline"
)

save_xlsx_list(
  list(pheno_continuous = pheno_cont, pheno_tertiles = pheno_tertiles),
  file.path(path_results, "tables", "ukbb_phenoage_main.xlsx")
)
```

## 3.2 KDM and early-onset cancer risk

```{r kdm-main-cox, eval=FALSE}
kdm_cont <- run_cox_series(
  data = dplyr::filter(ukbb_systemic_main, !is.na(Standardized_AA_kdm)),
  event_vars = outcomes$eo_main_figure,
  time_stop_var = "t_AllCancerEO55_age",
  exposure = "Standardized_AA_kdm",
  covariates = ukbb_covars,
  time_start_var = "ageBaseline"
)

kdm_cont <- add_bh_fdr(kdm_cont, p_col = "p.value")

save_xlsx_list(
  list(kdm_continuous = kdm_cont),
  file.path(path_results, "tables", "ukbb_kdm_main.xlsx")
)
```

# 4. Main Cox models: metabolomic aging

```{r metabolomic-main-cox, eval=FALSE}
# Recommended column names to standardize upstream:
# - MetaboAgeMort_calibrated
# - MetaboAgeMortAccel
# - Standardized_AA_MetaboAgeMortAccel

met_cont <- run_cox_series(
  data = dplyr::filter(ukbb_met_ready, !is.na(Standardized_AA_MetaboAgeMortAccel)),
  event_vars = outcomes$eo_main_figure,
  time_stop_var = "t_AllCancerEO55_age",
  exposure = "Standardized_AA_MetaboAgeMortAccel",
  covariates = ukbb_covars,
  time_start_var = "ageBaseline"
)

met_cont <- add_bh_fdr(met_cont, p_col = "p.value")

save_xlsx_list(
  list(metabolomic_continuous = met_cont),
  file.path(path_results, "tables", "ukbb_metabolomic_main.xlsx")
)
```

# 5. Main Cox models: organ-specific aging

```{r organ-main-cox, eval=FALSE}
organ_vars <- get_organ_clock_vars()

organ_results <- purrr::map_dfr(organ_vars, function(organ_var) {
  run_cox_series(
    data = dplyr::filter(ukbb_organ_ready, !is.na(.data[[organ_var]])),
    event_vars = c("dt_LungEO55", "dt_ColorectalEO55"),
    time_stop_var = "t_AllCancerEO55_age",
    exposure = organ_var,
    covariates = ukbb_covars,
    time_start_var = "ageBaseline"
  ) %>%
    mutate(organ = organ_var)
})

organ_results <- add_bh_fdr(organ_results, p_col = "p.value", group_cols = c("outcome"))

save_xlsx_list(
  list(organ_results = organ_results),
  file.path(path_results, "tables", "ukbb_organ_main.xlsx")
)
```

# 6. All of Us validation analyses

```{r aou-validation, eval=FALSE}
## Not publishable due to All of Us policy
```