---
title: "03 sensitivity analysis"
author: "Ruiyi Tian"
output: html_document
---

```{r setup, include=FALSE}
rm(list = ls())
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
source("R/functions.R")
load_core_packages()
```

# Purpose

This file contains analyses that support robustness checks and extended results, separated from the main analysis file:

1. exclude follow-up under 2 years
2. redefine early-onset cancer using `<50 years`
3. additional adjustment for telomere length and PRS
4. organ aging additionally adjusted for PhenoAge
5. repeated-measure stability analyses
6. optional mediation analyses

# 1. Load derived datasets

```{r load-data}
load("/path/to/Project/Data/Preprocessed_Project_Data_Derived.RData") 
```

# 2. Exclude follow-up under 2 years

```{r followup-ge2y, eval=FALSE}
ukbb_ge2y <- ukbb_systemic_main %>% filter(t_AllCancerEO55 >= 2 | dt_AllSolidCancerEO55 == 0)

pheno_ge2y <- run_cox_series(
  data = dplyr::filter(ukbb_ge2y, !is.na(Standardized_AA_phenoage)),
  event_vars = outcomes$eo_main_figure,
  time_stop_var = "t_AllCancerEO55_age",
  exposure = "Standardized_AA_phenoage",
  covariates = ukbb_covars,
  time_start_var = "ageBaseline"
)

save_xlsx_list(
  list(ukbb_pheno_followup_ge2y = pheno_ge2y),
  file.path(path_results, "tables", "sensitivity_followup_ge2y.xlsx")
)
```

# 3. Redefine early-onset cancer as diagnosis before age 50

```{r eo50, eval=FALSE}
# Recommended upstream naming convention:
# dt_LungEO50, dt_GastrointestinalEO50, dt_ColorectalEO50, ...
# t_AllCancerEO50_age

eo50_outcomes <- c(
  "dt_AllSolidCancerEO50", "dt_LungEO50", "dt_GastrointestinalEO50",
  "dt_ColorectalEO50", "dt_GIexcludeCRCEO50", "dt_UterusEO50"
)

pheno_eo50 <- run_cox_series(
  data = dplyr::filter(ukbb_systemic_main, !is.na(Standardized_AA_phenoage)),
  event_vars = eo50_outcomes,
  time_stop_var = "t_AllCancerEO50_age",
  exposure = "Standardized_AA_phenoage",
  covariates = ukbb_covars,
  time_start_var = "ageBaseline"
)

save_xlsx_list(
  list(ukbb_pheno_eo50 = pheno_eo50),
  file.path(path_results, "tables", "sensitivity_eo50.xlsx")
)
```

# 4. Additional adjustment for telomere length and PRS

```{r telomere-prs, eval=FALSE}
extra_covars <- c(
  ukbb_covars,
  "zadjusted_ts_log_f22192_0_0",  # telomere length
  "aging_prs_longevity",
  "aging_prs_lifespan",
  "lung_prs",
  "crc_prs",
  "endometrial_prs"
)

pheno_tel_prs <- run_cox_series(
  data = dplyr::filter(ukbb_systemic_main, !is.na(Standardized_AA_phenoage)),
  event_vars = c("dt_LungEO55", "dt_GastrointestinalEO55", "dt_UterusEO55"),
  time_stop_var = "t_AllCancerEO55_age",
  exposure = "Standardized_AA_phenoage",
  covariates = extra_covars,
  time_start_var = "ageBaseline"
)

save_xlsx_list(
  list(ukbb_pheno_telomere_prs = pheno_tel_prs),
  file.path(path_results, "tables", "sensitivity_telomere_prs.xlsx")
)
```

# 5. Organ aging additionally adjusted for systemic PhenoAge

```{r organ-plus-pheno, eval=FALSE}
organ_vars <- get_organ_clock_vars()

organ_adj_pheno <- purrr::map_dfr(organ_vars, function(organ_var) {
  run_cox_series(
    data = dplyr::filter(ukbb_organ_ready, !is.na(.data[[organ_var]]) & !is.na(Standardized_AA_phenoage)),
    event_vars = c("dt_LungEO55", "dt_ColorectalEO55"),
    time_stop_var = "t_AllCancerEO55_age",
    exposure = organ_var,
    covariates = c(ukbb_covars, "Standardized_AA_phenoage"),
    time_start_var = "ageBaseline"
  ) %>%
    mutate(organ = organ_var)
})

organ_adj_pheno <- add_bh_fdr(organ_adj_pheno, p_col = "p.value", group_cols = c("outcome"))

save_xlsx_list(
  list(organ_adjusted_for_phenoage = organ_adj_pheno),
  file.path(path_results, "tables", "sensitivity_organ_plus_pheno.xlsx")
)
```

# 6. Late-onset comparison models

```{r late-onset, eval=FALSE}
pheno_lo55 <- run_cox_series(
  data = dplyr::filter(ukbb_systemic_main, !is.na(Standardized_AA_phenoage)),
  event_vars = outcomes$lo_selected,
  time_stop_var = "t_AllCancerLO55_age",
  exposure = "Standardized_AA_phenoage",
  covariates = ukbb_covars,
  time_start_var = "ageBaseline"
)

pheno_lo55 <- add_bh_fdr(pheno_lo55, p_col = "p.value")

save_xlsx_list(
  list(ukbb_pheno_late_onset = pheno_lo55),
  file.path(path_results, "tables", "sensitivity_late_onset.xlsx")
)
```

# 7. Repeated-measure stability of PhenoAge tertiles

```{r repeated-measures, eval=FALSE}
# Expected upstream object: repeated biomarker dataset with baseline and repeat PhenoAge AA.

# repeated_df <- readRDS(file.path(path_derived, "ukbb_repeat_phenoage.rds"))
# repeated_df <- repeated_df %>%
#   mutate(
#     baseline_tert = add_quantile_group(AA_phenoage_baseline),
#     repeat_tert   = add_quantile_group(AA_phenoage_repeat)
#   )
#
# stability_tab <- repeated_df %>%
#   count(baseline_tert, repeat_tert) %>%
#   group_by(baseline_tert) %>%
#   mutate(prop = n / sum(n))
#
# saveRDS(stability_tab, file.path(path_results, "tables", "repeated_measure_stability.rds"))
```